---
title: "HW 3"
author: "Kelsey Blackstone"
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: inline
---
```{r, echo = FALSE, results='hide'}
library(kableExtra)
library(bivariate)
library(plotly)
library(invgamma)
library(coda)
```

## Problem 6: Confidence Intervals 

```{r, echo = FALSE, results='hide'}
n = 10
x = 1

# Exact
alpha_post = x + 1/2
beta_post = n - x + 1/2
exact_s1 = c(qbeta(.025, alpha_post, beta_post),qbeta(.975, alpha_post, beta_post))

e_1 = paste("(", toString(round(c(exact_s1[1], exact_s1[2]), 4)), ")")

# Laplace Approximation
theta0 <- (alpha_post - 1)/(alpha_post + beta_post - 2)

q0 <- exp((alpha_post - 1)*log(theta0) + (beta_post - 1)*log(1 - theta0))
A_inv <- (alpha_post - 1)*(beta_post - 1)/(alpha_post + beta_post - 2)^3

# Constant
k <- q0*sqrt(2*pi*A_inv)/beta(alpha_post, beta_post)

# Interval
lower <- qnorm((1 - 0.95/k)/2, theta0, sqrt(A_inv))
upper <- qnorm((1 - 0.95/k)/2, theta0, sqrt(A_inv), lower.tail=FALSE)
LA_s1 = paste("(", toString(round(c(lower, upper),4)), ")")

# Monte Carlo Simulated Interval
mc_sim <- rbeta(40000, alpha_post, beta_post)
MC_s1 = paste("(", toString(round(quantile(mc_sim, probs=c(0.025, 0.975)), 4)), ")")
```

```{r, echo=FALSE, results='hide'}
n = 100
x = 10

# Exact Interval
alpha_post = x + 1/2
beta_post = n - x + 1/2
exact_s2 = c(qbeta(.025, alpha_post, beta_post),qbeta(.975, alpha_post, beta_post))
e_2 = paste("(", toString(round(c(exact_s2[1], exact_s2[2]), 4)), ")")

# Laplace Approximation
theta0 <- (alpha_post - 1)/(alpha_post + beta_post - 2)

q0 <- exp((alpha_post - 1)*log(theta0) + (beta_post - 1)*log(1 - theta0))
A_inv <- (alpha_post - 1)*(beta_post - 1)/(alpha_post + beta_post - 2)^3

# Constant
k <- q0*sqrt(2*pi*A_inv)/beta(alpha_post, beta_post)

# Interval
lower <- qnorm((1 - 0.95/k)/2, theta0, sqrt(A_inv))
upper <- qnorm((1 - 0.95/k)/2, theta0, sqrt(A_inv), lower.tail=FALSE)
LA_s2 = paste("(", toString(round(c(lower, upper), 4)), ")")

# Monte Carlo Simulated Interval
mc_sim <- rbeta(40000, alpha_post, beta_post)
MC_s2 = paste("(", toString(round(quantile(mc_sim, probs=c(0.025, 0.975)),4)), ")")
```

```{r, echo=FALSE, results='hide'}
n = 1000
x = 100

# Exact
alpha_post = x + 1/2
beta_post = n - x + 1/2
exact_s3 = c(qbeta(.025, alpha_post, beta_post),qbeta(.975, alpha_post, beta_post))
e_3 = paste("(", toString(round(c(exact_s3[1], exact_s3[2]),4)), ")")

# Laplace Approximation
theta0 <- (alpha_post - 1)/(alpha_post + beta_post - 2)
theta0

q0 <- exp((alpha_post - 1)*log(theta0) + (beta_post - 1)*log(1 - theta0))
A_inv <- (alpha_post - 1)*(beta_post - 1)/(alpha_post + beta_post - 2)^3

# Constant
k <- q0*sqrt(2*pi*A_inv)/beta(alpha_post, beta_post)
k

# Interval
lower <- qnorm((1 - 0.95/k)/2, theta0, sqrt(A_inv))

upper <- qnorm((1 - 0.95/k)/2, theta0, sqrt(A_inv), lower.tail=FALSE)

LA_s3 = paste("(", toString(round(c(lower, upper),4)), ")") 

# Monte Carlo Simulated Interval
mc_sim <- rbeta(40000, alpha_post, beta_post)
MC_s3 = paste("(", toString(round(quantile(mc_sim, probs=c(0.025, 0.975)),4)), ")")

```

```{r, echo = FALSE}
x = c("Exact", "Laplace Approximation", "Monte Carlo Simulation")
y = c(e_1, LA_s1, MC_s1)
z = c(e_2, LA_s2, MC_s2)
m = c(e_3, LA_s3, MC_s3)
table = as.data.frame(cbind(x,y,z,m))
table %>% kable(caption = "Confidence Intervals for Different Approximations", col.names = c("CI Method", "n =  10, x = 1", "n = 100, x = 10", "n = 1000, x = 100")) %>% 
  kable_styling(bootstrap_options = c("striped", "hover"),
                latex_options = "HOLD_position")

```


## Problem 7 

### Part A: 
Find the posterior. Use the simulated dataset and evaluate the posterior density in the grid of ($\alpha$, $\beta$)

```{r}
## Generating 500 samples from the Gumbel Type II Distribution

set.seed(666)

a = 5
b = 5

u = runif(500)
x = (-log(u)/b)^(-1/a)

## Likelihood f(x | a, b):

loglike_gumb <- function (a, b) {500*log(a*b)+(-a-1)*sum(log(x))-b*sum(x^-a)}

## We now need to take the double integral with respect to a and b 
## in order to obtain the normalizing constant for the denominator

# for all combinations of a and b - multiply to get area
# store values into a matrix 

a = seq(4, 7, by=0.01)
b = seq(4, 7, by=0.01)
base=0.01*0.01

rectangles <- matrix(NA, nrow = length(a), ncol = length(b))
height <- matrix(NA,nrow = length(a), ncol = length(b))
for (i in 1:length(a)){
  for (j in 1:length(b)){
    result = exp(loglike_gumb(a[i], b[j]))
    height[i,j] = result
    rectangles[i,j] = base*result
  }
}

## Normalizing Constant
sum(rectangles)

## Final Posterior 
post_gumb = height/sum(rectangles)

## Plotting final posterior
layout(add_surface(plot_ly(x = a, y = b, z = post_gumb)),scene = list(xaxis = list(title = "alpha"), yaxis = list(title = "beta"), zaxis = list(title = "Posterior Gumbel Distribution")))

```

### Part B: 

Find the Laplace approximation to the posterior. Use the same simulated data and evaluate the approximated posterior on the grid of ($\alpha$, $\beta$).

```{r}
## Find Laplace Approximation to the normal

## Set posterior equal to q0

q0 = function(theta){
  a = theta[1]
  b = theta[2]
  500*log(a*b)+(-a-1)*sum(log(x))-b*sum(x^-a)}

# find the posterior mode (th0) and covariance matrix (A)

## we want to optimize alpha and beta
begin = c(a = 0.5, b = 0.5)

## make theta a vector of parameters in order to use optim function
## maximize by setting control to negative

LA_values = optim(begin, q0, control = list(fnscale = -1), hessian = TRUE)
th0 = LA_values$par             
cov_mat_A = LA_values$hessian 
cov_mat_Ainv = solve(-cov_mat_A)
th0
cov_mat_Ainv

# Density of Laplace Approximation to Normal

# variance of alpha = var.X is in cov_mat_Ainv[1,1]
# variance of beta = var.Y is cov_mat_Ainv[2,2]
# cov of a, b = cov is either cov_mat_Ainv[1,2] or cov_mat_Ainv[2,1]
# take multivariate normal and turn it into pdf 

LA_approx = nbvpdf.2(th0[1],th0[2],var.X = cov_mat_Ainv[1,1], 
                     var.Y = cov_mat_Ainv[2,2], cov = cov_mat_Ainv[1,2])
plot(LA_approx, TRUE, main="Laplace Approximation of Posterior",xlab="alpha",ylab="beta",zlab="density")


# compare post_gumb and LA_approx
a = seq(4, 7, by=0.01)
b = seq(4, 7, by=0.01)

height.la = matrix(NA, nrow = length(a), ncol = length(b))
for(i in 1:length(a)){
  for(j in 1:length(b)){
    height.la[i,j] = LA_approx(a[i],b[j])
  }
}

comp = height.la - height/sum(rectangles)

layout(add_surface(plot_ly(x = a, y = b, z = comp)),scene = list(xaxis = list(title = "alpha"), yaxis = list(title = "beta"), zaxis = list(title = "Laplace - Exact")))

```

### Part C: Comparing Exact and Laplace

Looking at the differences plot for the Laplace Approximated values vs. the exact values, we see that the differences are between 0.1 and -0.1, which isn't a large difference. This demonstrates that the Laplace Approximation to the normal, for large values of n (in this case, n=500), comes fairly close to the true values. 


## Problem 8

```{r}
# Problem 8
rm(list=ls())
set.seed(666)
n = 1000
X.st.dv = 1
X.mean = 5
data = rnorm(n, mean = X.mean, sd = X.st.dv)

# Setting values for hyperparamters for each case: (i), (ii), (iii), (vi)
a = c(1001, 0.1, 1001, 0.1)
b = c(1000, 0.1, 1000, 0.1)
k0 = c(0.01, 0.01, 100, 100)
theta0 = c(5, 5, 5, 5)
```


```{r, fig.height=3, fig.width=6}

par(mfrow = c(1,2))
# Part E: 

# (i) Fairly Informative Priors

# parameters of sigma^2|xis
shape_sigma_i = n/2 + a[1]
rate_sigma_i = ((X.st.dv)^2)*(n-1)/2 + (n*(X.mean - theta0[1])^2)/(2*(n*k0[1] + 1)) + b[1]

# parameters of theta|sigma, xi
theta0.mean_i = (n + 1/k0[1])^-1 * (n*X.mean + theta0[1]/k0[1])
theta0.var_i = 1/(n/(X.st.dv^2) + 1/(k0[1]*X.st.dv^2))

plot(density(rinvgamma(n, shape = a[1], rate = b[1])), main = "Informative Prior: Sigma^2")
plot(density(rinvgamma(n, shape_sigma_i, rate = rate_sigma_i)), main = "Posterior: Sigma^2")

plot(density(rnorm(n, mean = theta0[1], sd = k0[1]*X.st.dv)), main = "Informative Prior: Theta")
plot(density(rnorm(n, mean = theta0.mean_i, sd = sqrt(theta0.var_i))), main = "Posterior: Theta")


# (ii) Informative on theta and vague on sigma^2 

# parameters of sigma^2|xis
shape_sigma_ii = n/2 + a[2]
rate_sigma_ii = ((X.st.dv)^2)*(n-1)/2 + (n*(X.mean - theta0[2])^2)/(2*(n*k0[2] + 1)) + b[2]

# parameters of theta|sigma, xi
theta0.mean_ii = (n + 1/k0[2])^-1 * (n*X.mean + theta0[2]/k0[2])
theta0.var_ii = 1/(n/(X.st.dv^2) + 1/(k0[2]*X.st.dv^2))

plot(density(rinvgamma(n = 500, shape = a[2], rate = b[2])), main = "Vague Prior: Sigma^2")
plot(density(rinvgamma(n, shape_sigma_ii, rate = rate_sigma_ii)), main = "Posterior: Sigma^2")

plot(density(rnorm(n, mean = theta0[2], sd = k0[2]*X.st.dv)), main = "Informative Prior: Theta")
plot(density(rnorm(n, mean = theta0.mean_ii, sd = sqrt(theta0.var_ii))), main = "Posterior: Theta")


# (iii) Informative on sigma^2 and vague on theta
# parameters of sigma^2|xis
shape_sigma_iii = n/2 + a[3]
rate_sigma_iii = ((X.st.dv)^2)*(n-1)/2 + (n*(X.mean - theta0[3])^2)/(2*(n*k0[3] + 1)) + b[3]

# parameters of theta|sigma, xi
theta0.mean_iii = (n + 1/k0[3])^-1 * (n*X.mean + theta0[3]/k0[3])
theta0.var_iii = 1/(n/(X.st.dv^2) + 1/(k0[3]*X.st.dv^2))

plot(density(rinvgamma(n = 500, shape = a[3], rate = b[3])), main = "Informative Prior: Sigma^2")
plot(density(rinvgamma(n, shape_sigma_iii, rate = rate_sigma_iii)), main = "Posterior: Sigma^2")

plot(density(rnorm(n, mean = theta0[3], sd = k0[3]*X.st.dv)), main = "Vague Prior on Theta")
plot(density(rnorm(n, mean = theta0.mean_iii, sd = sqrt(theta0.var_iii))), main = "Posterior: Theta")


# (vi) Vague on sigma^2 and vague on theta
# parameters of sigma^2|xis
shape_sigma_vi = n/2 + a[4]
rate_sigma_vi = ((X.st.dv)^2)*(n-1)/2 + (n*(X.mean - theta0[4])^2)/(2*(n*k0[4] + 1)) + b[4]

# parameters of theta|sigma, xi
theta0.mean_vi = (n + 1/k0[4])^-1 * (n*X.mean + theta0[4]/k0[4])
theta0.var_vi = 1/(n/(X.st.dv^2) + 1/(k0[4]*X.st.dv^2))

plot(density(rinvgamma(n = 500, shape = a[4], rate = b[4])), main = "Vague Prior: Sigma^2")
plot(density(rinvgamma(n, shape_sigma_vi, rate = rate_sigma_vi)), main = "Posterior: Sigma^2")

plot(density(rnorm(n, mean = theta0[4], sd = k0[4]*X.st.dv)), main = "Vague Prior: Theta")
plot(density(rnorm(n, mean = theta0.mean_iii, sd = sqrt(theta0.var_iii))), main = "Posterior: Theta")

```

```{r}
## Part F

init = c(4,4) ## th0 and sigma0^2

niter = 10000
eta <- matrix(NA, nrow = niter, ncol = 4)   ## create empty matrix to store eta

# parameters of sigma^2 | xis
shape_sigma = n/2 + a
rate_sigma = ((X.st.dv)^2)*(n-1)/2 + 
  (n*(X.mean - theta0)^2)/(2*(n*k0 + 1)) + b

# parameters of theta | sigma, xi
theta0.mean = (n + 1/k0)^-1 * (n*X.mean + theta0/k0)
theta0.var = 1/(n/(X.st.dv^2) + 1/(k0*X.st.dv^2))

for (i in 1:niter){
    for (j in 1:4){
          # sampling new values for sigma and theta
          init[2] = rinvgamma(1, shape = shape_sigma[j], rate = rate_sigma[j])
          init[1] = rnorm(1, mean = theta0.mean[j], sqrt(init[2])) # new theta sample
          
          eta.value = init[1]/sqrt(init[2])
          eta[i,j] = eta.value
      }
}
```

```{r, echo=FALSE}
hist(eta[,1], main = "Case (i)")
hist(eta[,2], main = "Case (ii)")
hist(eta[,3], main = "Case (iii)")
hist(eta[,4], main = "Case (vi)")

## Example Plot 
plot(eta[,1],type="l", main = "Eta: Case (i)")
abline(5,0,lwd=2,col="red")

plot(eta[,2],type="l", main = "Eta: Case (i)")
abline(5,0,lwd=2,col="red")

plot(eta[,3],type="l", main = "Eta: Case (i)")
abline(5,0,lwd=2,col="red")

plot(eta[,4],type="l", main = "Eta: Case (i)")
abline(5,0,lwd=2,col="red")

## posterior means of eta
post.mean.eta1 <- mean(eta[seq(from=1001, to=niter,by=1),1]) 
post.mean.eta2 <- mean(eta[seq(from=1001, to=niter,by=1),2])
post.mean.eta3 <- mean(eta[seq(from=1001, to=niter,by=1),3])
post.mean.eta4 <- mean(eta[seq(from=1001, to=niter,by=1),4])

# Posterior 95% CI

quant1.low <- round(quantile(eta[seq(from=1001, to=niter, by=1), 1], 0.025),3)
quant1.upp <- round(quantile(eta[seq(from=1001, to=niter, by=1), 1], 0.975),3)
quant1 <- c(quant1.low, quant1.upp)

quant2.low <- round(quantile(eta[seq(from=1001, to=niter, by=1), 2], 0.025),3)
quant2.upp <- round(quantile(eta[seq(from=1001, to=niter, by=1), 2], 0.975),3)
quant2 <- c(quant2.low, quant2.upp)

quant3.low <- round(quantile(eta[seq(from=1001, to=niter, by=1), 3], 0.025),3)
quant3.upp <- round(quantile(eta[seq(from=1001, to=niter, by=1), 3], 0.975),3)
quant3 <- c(quant3.low, quant3.upp)

quant4.low <- round(quantile(eta[seq(from=1001, to=niter, by=1), 4], 0.025), 3)
quant4.upp <- round(quantile(eta[seq(from=1001, to=niter, by=1), 4], 0.975), 3)
quant4 <- c(quant4.low, quant4.upp)

y = c("i", "ii", "iii", "vi")
thet0 = c(5, 5, 5, 5)
k0 = c(0.01, 0.01, 100, 100)
a = c(1001, 0.1, 1001, 0.1)
b = c(1000, 0.1, 1000, 0.1)
post.mean = c(round(post.mean.eta1,3), round(post.mean.eta2,3), round(post.mean.eta3,3), round(post.mean.eta4,3))
cis_low = c(quant1.low, quant2.low, quant3.low, quant4.low)
cis_upp = c(quant1.upp, quant2.upp, quant3.upp, quant4.upp)
table = as.data.frame(cbind(y, thet0, k0, a, b, post.mean, cis_low, cis_upp))
table %>% kable(caption = "Posterior Quantities with various Prior Specification", col.names = c("Case", "θ", "k0", "a", "b", "Posterior Mean", "Posterior 2.5% CI", "Posterior 97.5% CI"), row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                latex_options = "HOLD_position")


```

